 \documentclass{article}
\begin{document}
This project is a comparative analysis of different existing methods of dealing with missing data in models of ecological time series. The attached manuscript is a draft that includes figures, complete introduction, methods, and results sections, and an outline of the discussion. We're eager to hear any and all general feedback, but in particular are interested in getting opinions on the following: 
\begin{enumerate}
    \item Is the level of detail consistent across the manuscript? Are there particular locations where detail is excessive or lacking? 
    \item We compared multiple different methods across multiple axes of variation in four different types of data, which is a lot to keep track of. Did we do a adequate job of explaining and keeping track of our methodology to the reader? 
    \item We've included a bulleted outline of the Discussion section (as well as rough drafts of intro and concluding paragraphs). Are there any conclusions that you drew that you think we should add (or focus more on) in the discussion? 
    \item Do you have any ideas for a target journal? We initally thought that Methods in Ecology and Evolution would be a good target, but we're worried about the fact that we don't present any \textit{new} methods, which is a stated requirement in MEE's submission guidelines (instead, we focus on the comparison of existing methods and adapting those methods to time series). The sparse modeling group's experience with their manuscript has also made us a bit nervous about finding the right journal. 
\end{enumerate}

Thanks so much for taking the time to read through this draft and provide comments! We really appreciate it. 

\end{document}